---
title: "Reproducible Research: Peer Assessment 1 by Muhammad Shahid"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r  echo=TRUE , message=FALSE}
library(data.table)
library(ggplot2)
library(lattice)
library(dplyr)
```

```{r echo=TRUE}
# Load the CSV RAW Data File
#setwd("D:\\RData\\Course5\\data")
df <- read.csv("activity.csv")
dt <- data.table(df)
dt$date<- as.Date(dt$date,"%Y-%m-%d")

```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day

```{r echo=TRUE}

steps_per_day <- dt[complete.cases(dt),] %>% 
  group_by(date) %>% 
  summarize("Total_Steps"=sum(steps))

```
2. Make a histogram of the total number of steps taken each day
```{r echo=TRUE}
hist(steps_per_day$Total_Steps , xlab = "Step" , ylab = "Frequency" , col="blue" , main="Total           Number of Steps taken Per day" , breaks=10 , xlim = c(0,22000),ylim = c(0,18) , labels=TRUE)
rug(steps_per_day$Total_Steps )
```

3. Calculate and report the mean and median of the total number of steps taken per day
  
```{r echo=TRUE}
        ## Calculate but save for use later
        mean_median <- 
                dt[complete.cases(dt),] %>% 
                        group_by(date) %>% 
                        summarize("Mean"=mean(steps),"Median"=median(steps))
        str(mean_median)
```

## What is the average daily activity pattern?
1. Make a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r echo=FALSE}
# Average daily Steps for each 5 minute interval
daily_activity <- dt[complete.cases(dt) , ] %>% group_by(interval) %>% summarize("Mean"= mean(steps))

plot( x=daily_activity$interval , y=daily_activity$Mean , type="l" , xlab = "Interval" , ylab="Mean of Total Steps" , main="Average Daily Activity" )

```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r echo=TRUE}
        x <- daily_activity[daily_activity$Mean==max(daily_activity$Mean),1]
        cat("Interval with Maximum mean of all days =",as.character(x))
```


## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs)

```{r echo=TRUE}
        x <- sum(!complete.cases(dt))
        cat("Number of rows with missing values =",as.character(x))
``` 

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r echo=TRUE}
print ("Fill the Missing Values with the mean of the interval of the day")
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r echo=TRUE}
        
        daily_activity <- as.data.frame(daily_activity)
        

        result <- dt

        ## Finding and Replacing Missing VALUES  - with Mean Values of the intervals
        for (i in 1:nrow(result)) {
                if (is.na(result$steps[i])) {
                        ## Lookup the mean (in daily_activity) for the interval of this row
                        val <- daily_activity[daily_activity$interval==result$interval[i],"Mean"]
                        result$steps[i]<- as.integer(round(val))
                }
        }
``` 

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day

```{r echo=TRUE}
        mean_median_imp <- 
                result %>% 
                group_by(date) %>% 
                summarize("Mean"=mean(steps),"Median"=median(steps))
        mean_median_imp
``` 

Analyze the difference  after Imputation of missing values
```{r echo=TRUE}
        summary(mean_median,2:3)
        summary(mean_median_imp,2:3)
``` 
Plot side by side plots before aqnd after imputation
```{r echo=TRUE}
        
        plot.new()
        par(mfrow=c(1,2) )
        
        step_summary <- dt[complete.cases(dt),] %>% 
                group_by(date) %>% 
                summarize("Total_Steps"=sum(steps))
        
        hist(step_summary$Total_Steps,col="blue",
             main="Avg. Daily Steps Before Imputation",
             xlab="Total Steps"
             )
        
        step_summary <-result %>%
                group_by(date) %>% 
                summarize("Total_Steps"=sum(steps))
        
        hist(step_summary$Total_Steps,col="green",
             main="Avg. Daily Steps After Imputation",
             xlab="Total Steps"
             )
``` 

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r echo=TRUE}
        
        DayOfWeek <- data.frame(
                name=c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"),
                part=c("weekday","weekday","weekday","weekday","weekday","weekend","weekend"))
       
        # Parse the Date to Names of the Week days and assign the FACTOR value from the Data Frame         of DaysOfWeek
        result$week_factor <- as.factor(weekdays(result$date))
        result$week_factor <- DayOfWeek[result$week_factor,2]
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r echo=TRUE}
        ## Calculate and store the means by group
        plot <- result %>% 
                group_by(week_factor,interval) %>% 
                summarize("Mean"=mean(steps))
        
        
        xyplot(plot$Mean ~ plot$interval | plot$week_factor, 
               data=plot, type="l", layout=c(1:2),
               ylab="Mean Number of Steps",
               xlab="Interval")
```
